Research Experience

### 2010-03-23 doc/thesis/thesis.tex
guh more on the particle filters, although I'm staring to \emph{really} understand

\subsection*{Introduction to Particle Filters}
Particle filters, also known as Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, particle
filters extend relatively easily to any finite impulse response (FIR) system.
Unlike Variational Bayes or simple gradient descent algorithms, particle
filters can be run online, with relatively lower cost. The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF obviously introduces bias; however, it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose either a set or stream
of measurements $y_t$ are made at discrete times $t = 0, 1, ... $, where
$y_t$ could be a vector of measurements. Then the goal is to find the set
of parameters $\hat{\theta}$ that give the best estimate, $\hat{y}_t$, of $y_t$.
In a biological case, with a model based on first principals, there is 
presumably some true set of parameters, $\theta$, a true model $f$
and a true observation equation, $g$ that dictate how the stimulus 
produce noisy measurements,
\begin{equation}
\dot{x}(t) = f(t, x(t), u(t), \theta) + \nu_d
\end{equation}

\begin{equation}
y(t) = g(t, x(t), u(t), \theta) + \nu_s
\end{equation}

Where $x(t)$ is a vector of state variables, $\theta$ is a vector of system
constants, $\nu_d$ and $\nu_s$ are bother stationary zero-mean distributions, $u(t)$
is timeseries of stimuli and
$y(t)$ is the vector of observations. $\nu_d$, which produces a drift term in $y(t)$
may be driven by unknown changes in 
cerebral blood flow, or blood oxygen content. $\nu_s$ will produce standard
stationary white noise in $y(t)$ and may encapsulate noise by unknown changes due 
to the heart beat. There are also significant non-biological effects that 
could drive these noise terms such as inherent MR noise.

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, that $\theta$ is time invariant. This assumption is 
based on the idea that most physiological parameters don't change in the 
short run unless they are driven by some other stimulus. The obvious exception
might be something like the heart; however, period between heartbeats vary quickly enough that accurately accounting for them would necessitate an entirely
new set of model parameters, or direct outside information. In short we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so ananticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Finally, $x(t)$ will include $\theta$, the unknown model constants, although
obviously that will mean that the vector $\dot{x}$ will always have some members
that are constant and 0. The state space equations are then simplified to:
\begin{equation}
\dot{x}(t) = f(x(t), u(t)) + \nu_d
\end{equation}

\begin{equation}
y(t) = g(x(t)) + \nu_s
\end{equation}

The goal of the particle filter, is then to derive the probably distribution
of the value of $x(t)$
for each time, given known inputs $u(t), u(t-1), ... u(0)$  and noise 
measurements produced by the system, $y(t), ... , y(0)$ (every time up to
$t$). To begin with then, the particle filter starts with a prior distribution,
which is used to build a Dirac Mixture Probability Distribution Function (PDF):

@@

Thus $\hat{x}(0)$ is an empirical representation of the prior, and $\hat{x}_i(0)$
is the $i^{th}$ particle, a single possible representation of the system at time 0.
Initially, since each $\hat{x}_i$ has been drawn from the prior, and no measurements
have been made, the particles are all equally weighted. When a measurement becomes
available, that information is then incorporated into the posterior by re-weighting
the particles according to how well they estimated this new measurement. 
\begin{equation}
Pr[\hat{x}_i(t) | y(0:t-1)] = \int Pr[\hat{x}_i(t) | \hat{x}_i(t-1), y(0:t-1)]Pr[\hat{x}_i(t-1)
\end{equation}
Since $\hat{x}(t) = \hat{x}(t-1) + \int_{t-1}^t f(x(t), u(t))dt + \int_{t-1}^t\nu_d dt$, 
\begin{equation}
Pr[\hat{x}_i(t) | \hat{x}_i(t-1)] = Pr[\int{t-1}^t \nu_d]
\end{equation}

>>>>

\subsection*{Introduction to Particle Filters}
Particle filters, a type of Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, Sequential Monte-Carlo methods are designed
to be used with parameters that vary with time. Unlike variations of the
Kalman filter, particle filters do not make the assumption that noise
is Gaussian. Thus particle filters are often the best solution to bayesian 
tracking for non-linear, non-gaussian systems. 
%todo: might need to reference attempts to model a system of our type with
%unscented kalman filter
The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF obviously introduces bias; however, every gradient
descent must choose starting points and it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose a set or stream of measurements
are given, $\{y(t), t = 1, 2, 3, ... T\}$, where $T$ is permitted to go
 to infinity. Then the goal is to find the 
parameters, $\hat{\theta}$, and underlying state time series, $\hat{x}[0:T]$
that minimize the difference between $\hat{y}[0:T]$
and $y[0:T]$. In our case, we will assume that we know the form
of the model, which is based on first principals, and that
there is some true $\theta$ and a true time-series of underlying
state variable, $x[0:T]$ that drives $y[0:T]$. Assuming a model form 
such as we do here reduces model variance, potentially at the cost of increased
bias (or systematic) error. We will assume a basic state space model:

\begin{equation}
\dot{x}(t) = f(t, x(t), u(t), \theta, \nu_d)
\end{equation}

\begin{equation}
y(t) = g(t, x(t), u(t), \theta, \nu_s)
\end{equation}

Where $x(t)$ is a vector of state variables, $\theta$ is a vector of system
constants, $u(t)$ is a stimulus, $y(t)$ an observation, and
$\nu_x$ and $\nu_y$ are random variates. Obviously any one of these could
be a vector, so for instance $u(t)$ could encode multiple types of stimuli.

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, the systems are assumed to be time invariant. This 
assumption is based on the idea that if you froze the system for $\delta t$
seconds, when unfrozen the system would continue as if nothing happend. 
Few biological systems are predictible enough for them to be summarized
by a time varying function. Although the heart may seem like an obvious
exception, period between heartbeats vary often enough that prediction
would necessate another state-space model. In short, we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Third, we will assume noise is additive.
Finally, $x(t)$ will encapsulate $\theta$, the unknown model constants, which
means that the vector $\dot{x}$ will always have members
that are 0. The results of these assumptions are a simplified version of the
state space equations:

\begin{equation}
\dot{x}(t) = f(x(t), u(t)) + \nu_x
\end{equation}

\begin{equation}
y(t) = g(x(t)) + \nu_y
\end{equation}

$\nu_x$ will result in something akin to adding a weiner process to $y[0:T]$, and thus
will include low frequency noise. $\nu_y$ on the other hand will cause i.i.d. noise
in $y[0:T]$. For some tests I will use de-trending methods to reduce the effects of 
$\nu_x$, the remainder of which gets re-factored into $\nu_y$. Both $\nu_x$ and $\nu_y$
have biological and non-biological sources. MR can lead to both types of noise, 
as demonstrated in \cite{drift}. Meanwhile changes in metabolism, heart rate, or
other biochemical intervention could all lead to either $\nu_x$ or $\nu_y$.

The goal of the particle filter to evolve a probably distribution $\hat{x}(T)$,
that asymptotically approaches the probability distribution of $x(t)$,
given known inputs, $u[0:T]$ and noisy readings $y[0:T]$.
To begin with, the particle filter starts with a prior distribution,
which is used to build a Dirac Mixture Probability Distribution Function (PDF):

@@

Thus $\hat{x}(0)$ is an empirical representation of the prior, and $\hat{x}_i(0)$
is the $i^{th}$ particle, a single possible representation of the system at time 0.
Initially, since each $\hat{x}_i$ has been drawn from the prior, and no measurements
have been made, the particles are all equally weighted. When a measurement becomes
available, that information is then incorporated into the posterior by re-weighting
the particles according to how well they estimated this new measurement.

The estimated posterior of $\hat{x}(t)$, for discrete time, $t$ (before incorporating
any measurement that might be available at $t$) is then:
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t-1]\} =  \int \Pr\{\hat{x}(t), \hat{x}(t-1) | y[0:t-1]\} d\hat{x}(t-1)
\end{equation}

and by bayes theorem $\Pr\{\hat{x}(t), \hat{x}(t-1) | y[0:t-1]\} = 
\Pr\{\hat{x}(t) | \hat{x}(t-1), y[0:t-1]\}\Pr\{\hat{x}(t-1) | y[0:t-1]\}$,
therefore 
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t-1]\} =  \int \Pr\{\hat{x}(t) | \hat{x}(t-1) ,y[0:t-1]\}\Pr\{\hat{x}(t-1) | y[0:t-1]\}  d\hat{x}(t-1)
\end{equation}

Since in this work we will only deal with markov processes, we assume that 
$\Pr\{\hat{x}(t) | \hat{x}(t-1), y[0:t-1]\} = \Pr\{\hat{x}(t) | \hat{x}(t-1)\}$

thus the posterior becomes:
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t-1]\} =  \int \Pr\{\hat{x}(t) | \hat{x}(t-1)\}\Pr\{\hat{x}(t-1) | y[0:t-1]\}  d\hat{x}(t-1)
\end{equation}

Therefore in between measurements, there is no need to know previous measurements in
order to make predictions about the internal state. This is important because it means makes
particle filters significantly more memory efficient, and although it isn't useful in this
case, it means a particle filter could theoretically be run indefinitely without the
need to store old data.

As stated previously, new measurements may be incorporated by reweighting particles.
To find the updated distribution, $\Pr\{\hat{x}(t) | y[0:t]\}$, bayes theorem is
again applied:
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t]\} = \Pr\{\hat{x}(t) | y[0:t-1], y(t)\} = \frac{\Pr\{y(t) | \hat{x}(t), y[0:t-1]\}\Pr\{\hat{x}(t) | y[0:t-1]\}}
{\Pr\{y(t) | y[0:t-1]\}}
\end{equation}

\begin{equation}
\Pr\{\hat{x}_i(t) | y[0:t-1]\} = \int \Pr\{\hat{x}_i(t) | \hat{x}_i(t-1), y[0:t-1]\}\Pr[\hat{x}_i(t-1)
\end{equation}
Since $\hat{x}(t) = \hat{x}(t-1) + \int_{t-1}^t f(x(t), u(t))dt + \int_{t-1}^t\nu_d dt$, 
\begin{equation}
\Pr\{\hat{x}_i(t) | \hat{x}_i(t-1)\} = \Pr\{\int{t-1}^t \nu_d\}
\end{equation}
